---
title: "World life expectancy analysis"
author: "Lara Komel"
date: "6/7/2021"
output: html_document
---

```{r setup, include=FALSE, message = FALSE, warning = FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
setwd("//Users//larakomel")
read.csv("Documents/4.semester/podatk prog/LaraKomelprojekt/life.csv")%>%as.tibble()->lifeDF1
#lifeDF%>%filter(Country!="Cook Islands")->lifeDF
lifeDF1%>%filter(!is.na(Life.expectancy))->lifeDF
```

## Introduction and presentation
[Life expectancy](https://ourworldindata.org/life-expectancy) is the key for assessing population health, and predicting the future health.
Broader than the narrow metric of the infant and child mortality, which focus solely at mortality at a young age, life expectancy captures the mortality along the entire life course. It tells us the average age of death in a population.

Estimates suggest that in a pre-modern, poor world, life expectancy was around 30 years in all regions of the world.
Since 1900 the global average life expectancy has more than doubled and is now above 70 years. The inequality of life expectancy is still very large across and within countries.

For my project I chose the Life Expectancy (WHO) data set to analyze. The file includes data for 193 different countries, and tracks life expectancy from the year 2000 to 2015, mortality in different age groups, diseases like Hepatitis, measles, HIV, etc.

All predicting variables were then divided into several broad categories: Immunization related factors, Mortality factors, Economical factors and Social factors.

## Presentation of the data
The dataset has the following 22 columns:

+ Country
+ Year
+ Status (Developed or Developing status)
+ Life Expectancy (in age)
+ Adult Mortality (Adult Mortality Rates of both sexes (probability of dying between 15 and 60 years per 1000 population))
+ Infant deaths (Number of Infant Deaths per 1000 population)
+ Alcohol (recorded per capita (15+) consumption (in litres of pure alcohol))
+ Percentage expenditure (Expenditure on health as a percentage of Gross Domestic Product per capita(%))
+ Hepatitis B (Hepatitis B (HepB) immunization coverage among 1-year-olds (%))
+ Measles (Measles - number of reported cases per 1000 population)
+ BMI (Average Body Mass Index of entire population)
+ under-five deaths (Number of under-five deaths per 1000 population)
+ Polio (Polio (Pol3) immunization coverage among 1-year-olds (%))
+ Total expenditure (General government expenditure on health as a percentage of total government expenditure (%))
+ Diphtheria (Diphtheria tetanus toxoid and pertussis (DTP3) immunization coverage among 1-year-olds (%))
+ HIV/AIDS (Deaths per 1 000 live births HIV/AIDS (0-4 years))
+ GDP (Gross Domestic Product per capita (in USD))
+ Population (Population of the country)
+ thinness 1-19 years (Prevalence of thinness among children and adolescents for Age 10 to 19 (% ))
+ thinness 5-9 years (Prevalence of thinness among children for Age 5 to 9(%))
+ Income composition of resources (Human Development Index in terms of income composition of resources (index ranging from 0 to 1))
+ Schooling (Number of years of Schooling(years))

```{r}
lifeDF%>%head()
lifeDF%>%colnames()
#lifeDF%>%summary()
```

__Number of rows that contain missing values:__
```{r}
which_nas <- apply(lifeDF1, 1, function(X) any(is.na(X)))
length(which(which_nas))
```

__First plots are created for a few columns:__
```{r fig1, fig.height = 3, fig.width = 5, warning = FALSE, message=FALSE}
ggplot(lifeDF, aes(x=Status)) + 
  geom_histogram(color="black", fill="white", stat="count")
lifeDF%>%group_by(Status)%>%summarise(Num = n())->status
ggplot(status, aes(x="", y=Num, fill=Status)) +
  geom_bar(stat="identity", width=1) +
  coord_polar("y", start=0)
ggplot(lifeDF, aes(x=Life.expectancy)) + 
  geom_histogram(color="black", fill="white")
```
  
From the graphs we can see that we are looking at mostly developing countries in the dataset, and that life expectancy around 75 appears most in the data, where as under 50 and over 80 is more rare.   


__Mean for a few columns relating to deaths:__
```{r}
colMeans(lifeDF%>%select(Life.expectancy, Adult.Mortality, infant.deaths, under.five.deaths))
```
The avarage life expectancy is 96 years, about 164 adults per 1000 people die between ages 15 and 60 on avarage, about 30 infants and 42 kids under 5.  
__Average life expectancy in the years 2000-2015 for every country:__
```{r}
lifeDF%>%group_by(Country)%>%summarise(Life.expectancy=mean(Life.expectancy))->life
life
life%>%top_n(3, Life.expectancy)
life%>%slice_min(Life.expectancy, n=3)
```
Top three countries with the highest avarage life expectancy are Iceland, Japan and Sweden, where the lowest are Sierra Leone, Central African Republic and Lesotho.  
  
__How much did life expectancy rise(or not) in the fifteen years for each country (of those which have the data):__
```{r}
raise<-function(country){
  min = (lifeDF%>%filter(Country==as.character(country) & Year==2000))$Life.expectancy
  max = (lifeDF%>%filter(Country==as.character(country) & Year==2015))$Life.expectancy
  max-min
}
lifeDF%>%group_by(Country)%>%mutate(Change = raise(Country))%>%summarise(Change = mean(Change))->change
change
change%>%top_n(3, Change)
change%>%slice_min(Change, n=3)
```

Countries with the biggest jump in life expectancy are Eritrea, Zambia and Zimbabwe, where as expectancy fell in Syrian Arav Republic, Saint Vincent and the Grenadines and Libya.  

__The avarage life expectancy for every year:__
```{r}
lifeDF%>%group_by(Year)%>%summarise(Life.expectancy=mean(Life.expectancy))->av
av
difference = av$Life.expectancy%>%max-av$Life.expectancy%>%min
```
The average world life expectancy raised by `r difference` in the years 2000 to 2015, and was the highest in 2015.  

__Countries with the lowest and highest life expectancy in 2000 are:__
```{r}
lifeDF%>%filter(Year==2000)%>%filter(Life.expectancy == min(Life.expectancy))%>%select(Country, Life.expectancy)
lifeDF%>%filter(Year==2000)%>%filter(Life.expectancy == max(Life.expectancy))%>%select(Country, Life.expectancy)
```
In 2000, the country with the highest expectancy was Japan, where the result is over twice as much as in Sierra Leone, where it was the lowest at only 39.  

__Countries with the lowest and highest life expectancy in 2015 are:__
```{r}
lifeDF%>%filter(Year==2015)%>%filter(Life.expectancy == min(Life.expectancy))%>%select(Country, Life.expectancy)
lifeDF%>%filter(Year==2015)%>%filter(Life.expectancy == max(Life.expectancy))%>%select(Country, Life.expectancy)
```
In 2015, the lowest was still in Sierra Leone, even though it did rise by 12 years, and was the highest is Slovenia, almost 7 years more than the highest age in 2000.  

__The avarage life expectancy, adult mortality and infant mortality for developed and undeveloped countries:__
```{r}
lifeDF%>%group_by(Status)%>%summarise(Life.expectancy=mean(Life.expectancy), Adult.Mortality=mean(Adult.Mortality), infant.deaths = mean(infant.deaths))
```
Also a fairly big difference is life expectancy and mortality between countries that are already developed and ones which are still developing. The avarage life expectancy is 12 years higher in the developed countries, and, where in the developed world only one infant dies on avarage per 1000 people, in the developing world the number is about 30 times higher.  

__Thinness in relation to income of countries:__
```{r fig5, fig.height = 3, fig.width = 5}
lifeDF$BMI = as.numeric(lifeDF$BMI)
lifeDF$thinness..1.19.years = as.numeric(lifeDF$thinness..1.19.years)
lifeDF%>%group_by(Country)%>%summarise(Thinness=mean(thinness..1.19.years), GDP = mean(GDP))%>%filter(!is.na(Thinness), !is.na(GDP))->df1
df1%>%ggplot(aes(x=Thinness, y=GDP))+geom_line()
```

Another relation is between the Gross Domestic Product per capita (in USD) of the country and how many percantage of children are underweight. Unsurprisingly, we can see that, with a few exceptions, the poorer citizens are, the bigger percentage of children in underweight.  

__Difference between life expectancy for developed and developing countries:__
```{r fig13, fig.height = 3, fig.width = 5}
lifeDF%>%group_by(Status)%>%summarise(Life.expectancy = mean(Life.expectancy))->stat1
ggplot(stat1, aes(x = Status, y = Life.expectancy)) +
       geom_bar(stat="identity")

```
  
  By comparing the life expectancy for developed and undeveloped world, we can see the difference of about 12 years.  
  
  __And the, a lot more noticible difference between infant deaths in developed and undeveloped countries:__
```{r fig32, fig.height = 3, fig.width = 5}
lifeDF%>%group_by(Status)%>%summarise(infant.deaths = mean(infant.deaths))->infants
ggplot(infants, aes(x = Status, y = infant.deaths)) +
       geom_bar(stat="identity")
```
  

__Life expectancy in Slovenia for given 15 years:__
```{r fig4, fig.height = 3, fig.width = 5}
lifeDF%>%filter(Country=="Slovenia")-> sl
ggplot(sl, aes(x=Year, y=Life.expectancy))+geom_point()

```
  
  Looking only at the data for Slovenia, we can see a steady climb in life expectancy over the years, with a lot more steap climb in the most recent few years.   

In conclusion, we can see that life expectancy has been steadily climbing in the 15 years the data was collected. There is a noticable difference between life expectancy in the developed and undeveloped world, but even bigger difference in number of infants per 1000 people that die, as well as number of adults that die between the ages 15 and 60. Unsurprisingly, the countries with the "best" results in each category are mostly western (mostly european) or Asian, and the ones with "worst" results are third world countries.  

__References:__

+ [Life expectancy](https://ourworldindata.org/life-expectancy)
+ [Data](https://www.kaggle.com/kumarajarshi/life-expectancy-who/code)
+ [Developing country](https://en.wikipedia.org/wiki/Developing_country)